National demographic-disturbance relationships and local monitoring both reduce uncertainty in projections of boreal caribou population viability: tools to support monitoring decisions

### 2023-07-26 analysis/paper/paper.Rmd
In progress - writeup of methods and results.

@@

All regression coefficients are assumed to be Gaussian distributed, with priors derived from the expected values and 95% confidence intervals estimated by Johnson et al (2020) (Table \@ref(tab:priors)). We calibrated the prior distributions so that the 95% prediction interval from the Bayesian model matches the range between the 2.5% and 97.5% quantiles of 1000 simulated recruitment (Appendix A).

@@

@@

## A method for simulating example local demographic data

### Population dynamics
To simulate plausible plausible demographic trajectories for example populations we used a slightly modified version of Johnson et al's [-@johnson_science_2020] two-stage demographic model described by Dyson et al. [-@dyson_existing_2022]. The true number of post-juvenile females that survive from year $t$ to the census $\dot{W}_t$ is binomially distributed with true survival probability $\dot{S}_t$: $\dot{W}_{t} \sim \text{Binomial}(\dot{N}_t,\dot{S}_t)$. Maximum potential recruitment rate is adjusted for sex ratio, misidentification biases (Appendix X) and (optionally) delayed age at first reproduction $$\dot{X}_t=\frac{c\dot{R}_t/2}{1+c\dot{R}_t/2};c=\frac{w(1+qu-u)}{(w+qu-u)(1-z)}, $$
where $w$ is a multiplier that defines the apparent number of adult females in the composition survey as a function of the number of collared animals, $q$ the ratio of young bulls to adult females, $u$ is the probability of misidentifying young bulls as adult females and vice versa, and $z$ is the probability of missing a calf.

Interannual variation in survival and recruitment is modelled using truncated beta distributions [rtrunc function; @novomestky_package_2016]: $\dot{R}_t \sim \text{TruncatedBeta}(\bar{R}_t,\nu_R,l_R,h_R); \dot{S}_t \sim \text{TruncatedBeta}(\bar{S}_t,\nu_S,l_S,h_S)$. Coefficients of variation among years $(\nu_R,\nu_S)$ and maximum/minimum values $l_R,h_R,l_S,h_S$ for recruitment and survival are given in Table X. Expected recruitment ($\bar{R}_t$) and survival ($\bar{S}_t$) vary with disturbance according the beta regression models estimated by Johnson et al. [@johnson_science_2020]: $$ \bar{R}_t \sim Beta(\mu^R_t,\phi^R);log(\mu^R_t)=\dot{\beta^R_0}+\dot{\beta^R_a}A_t+\dot{\beta}^R_fF_t,$$
$$\bar{S}_t \sim (46\times Beta(\mu^S_t,\phi^S)-0.5)/45;log(\mu^S_t)=\dot{\beta^S_0}+\dot{\beta^S_a}A_t.$$
$\phi^R \sim \text{Normal}(19.862,2.229)$ and $\phi^S \sim \text{Normal}(63.733,8.311)$ are precisions of the Beta distributed errors [@ferrari_beta_2004]. At the beginning of a simulation for an example population, regression coefficient values are sampled from Gaussian distributions (see Table \@ref(tab:priors)) and the population is assigned to quantiles of the Beta error distributions for survival and recruitment. The population remains in these quantiles as disturbance changes over time, so there is substantial persistent variation in recruitment and survival among example among populations (example figure).

## Monitoring scenarios

We assume collars are deployed in January, and last for up to 6 years. The target number of collared females $n$ varies among scenarios, as does the total duration of the monitoring program (Table \@ref(tab:monitoringScenarios)). Collars lost to mortality may be replaced each year ($o=1$), or there may be a 3 year gap between deployments ($o=4$).

In our simulated populations, mortality risk does not vary over time or among individuals, so the observed state (dead or alive) of a collared animal $i$ in month $m$ and year $t$ depends on the true survival rate $\dot{S}_t$ (eq X) as follows: 
$$\hat{I}_{i,t,m} \sim \text{Bernoulli}(\dot{S}^{1/12}_t\hat{I}_{i,t,m-1}).$$
### Observed recruitment from composition surveys

## Disturbance scenarios

```{r fig1, fig.cap="Anthropogenic disturbance and monitoring scenarios.", cache=TRUE}
knitr::include_graphics(here::here("figs/forPaper/distScnscmult3ay3aSf1.png"))

@@

TO DO: consider showing results in terms of average number of collars deployed per year rather than target # and renewal interval.

```{r scn1, cache=TRUE,echo=FALSE, message=FALSE,results='hide'}
if(runScenarios){
  numObsYrs=c(2);startsByYr = 30 #25
  scns=expand.grid(P=numObsYrs,sQ=c(0.5),st=startsByYr)
  scResults = runScenario(scns)

  png(here::here("figs/Figure1_Example.png"),
    height = 4, width = fig_widths["mid"], units = "in",res=600)
  print(plotRes(scResults$rr.summary.all, "Female population size",obs=scResults$obs.all,
              lowBound=0,highBound=2000,facetVars=c("P","sQ")))
  dev.off()
}

```{r fig2, fig.cap="Example projections, assuming a target of 30 collars, renewed every 4 years, with 3 cows per collared cow included in aerial surveys, and a 1% increase in anthropogenic disturbance per year (as in Figure 1). The number of monitoring years varies from P=0 (top row) to P=24 (bottom row), and the anthropogenic disturbance footprint in 2023 varies from 0% (left column) to 60% (right column). A status assessment error occurs when the projected population growth rate is above one and the true rate is below one or vice versa. When disturbance is very high (e.g. Anthro2023 = 60) populations that differ substantially from one another (compare 'true' lines for high and low quantile populations) share the same status, and a projection informed only by national demographic disturbance relationships (P=0) is accurate for both. When disturbance is intermediate, populations that differ substantially from one another are more likely to differ in their status. In these cases, integrating local monitoring data improves projection accuracy (i.e. smaller distances between projected and true lines) and reduces status assessment errors.", cache=TRUE}
knitr::include_graphics(here::here("figs/forPaper/examples.png"))
```

```{r fig3, fig.cap="Power analysis example result.", cache=TRUE}
knitr::include_graphics(here::here("figs/forPaper/powercmult3ay3aSf1.png"))
knitr::include_graphics(here::here("figs/forPaper/powerEffortcmult3ay3aSf1.png"))

#TO DO: figure out how to put no monitoring results on these graphs.
```

@@

cr=expand.grid(w = c(1,2,3,4,5,6,7,8,9),rep=seq(1,10000))
nr=nrow(cr)
cr$c = compositionBiasCorrection(q=runif(nr,0,0.6),w=cr$w,u=runif(nr,0,0.2),z=runif(nr,0,0.2))

#get mean and sd as fn of w
cs <- cr %>%
  group_by(w) %>%
  summarise(mean = mean(log(c)), sd = sd(log(c)))

cr = merge(cr,cs)
cr$ca= exp(rnorm(nrow(cr),cr$mean,cr$sd))

>>>>

## Anthropogenic disturbance and monitoring scenarios

```{r figDisturbance, fig.cap="Anthropogenic disturbance and monitoring scenarios.", cache=TRUE}
knitr::include_graphics(here::here("figs/forPaper/distScnscmult6ay3aSf1.png"))

For this investigation we assume collars are deployed in January, and last for up to 6 years. The target number of collared females $n$ varies among scenarios, as does the total duration of the monitoring program (Table \@ref(tab:monitoringScenarios)). Collars lost to mortality may be replaced each year ($o=1$), or there may be a 3 year gap between deployments ($o=4$).

## Simulation of local population dynamics and monitoring

### Population dynamics informed by national disturbance-demographic relationships

```{r exampleTrajectories, fig.cap="Examples of simulated population trajectories from a model informed only by national disturbance-demographic relationships with (top row) and without (bottom row) delayed reproduction. White dots show means of 500 example trajectories. For reference, black lines and and shaded bands show means and 95\\% predictive intervals from Johnson et al's (2020) regression models without adjustment for misidentification bias and delayed reproduction.", cache=TRUE}

knitr::include_graphics(here::here("analysis/paper/figs/DemographicRatesDelay.png"))

To simulate plausible plausible demographic trajectories for example populations we used a modified version of Johnson et al's [-@johnson_science_2020] two-stage demographic model described by Dyson et al. [-@dyson_existing_2022]. The true number of post-juvenile females that survive from year $t$ to the census $\dot{W}_t$ is binomially distributed with true survival probability $\dot{S}_t$: $\dot{W}_{t} \sim \text{Binomial}(\dot{N}_t,\dot{S}_t)$. Maximum potential recruitment rate is adjusted for sex ratio, misidentification biases (Appendix X) and (optionally) delayed age at first reproduction $$\dot{X}_t=\frac{c\dot{R}_t/2}{1+c\dot{R}_t/2};c=\frac{w(1+qu-u)}{(w+qu-u)(1-z)}, $$
where $w$ is a multiplier that defines the apparent number of adult females in the composition survey as a function of the number of collared animals, $q$ the ratio of young bulls to adult females, $u$ is the probability of misidentifying young bulls as adult females and vice versa, and $z$ is the probability of missing a calf.

Interannual variation in survival and recruitment is modelled using truncated beta distributions [rtrunc function; @novomestky_package_2016]: $\dot{R}_t \sim \text{TruncatedBeta}(\bar{R}_t,\nu_R,l_R,h_R); \dot{S}_t \sim \text{TruncatedBeta}(\bar{S}_t,\nu_S,l_S,h_S)$. Coefficients of variation among years $(\nu_R,\nu_S)$ and maximum/minimum values $l_R,h_R,l_S,h_S$ for recruitment and survival are given in Table X. Expected recruitment ($\bar{R}_t$) and survival ($\bar{S}_t$) vary with disturbance according the beta regression models estimated by Johnson et al. [@johnson_science_2020]: $$ \bar{R}_t \sim Beta(\mu^R_t,\phi^R);log(\mu^R_t)=\dot{\beta^R_0}+\dot{\beta^R_a}A_t+\dot{\beta}^R_fF_t,$$
$$\bar{S}_t \sim (46\times Beta(\mu^S_t,\phi^S)-0.5)/45;log(\mu^S_t)=\dot{\beta^S_0}+\dot{\beta^S_a}A_t.$$
$\phi^R \sim \text{Normal}(19.862,2.229)$ and $\phi^S \sim \text{Normal}(63.733,8.311)$ are precisions of the Beta distributed errors [@ferrari_beta_2004]. At the beginning of a simulation for an example population, regression coefficient values are sampled from Gaussian distributions (see Table \@ref(tab:priors)) and the population is assigned to quantiles of the Beta error distributions for survival and recruitment. The population remains in these quantiles as disturbance changes over time, so there is substantial persistent variation in recruitment and survival among example among populations (Figure \@ref(fig:exampleTrajectories)).

In our simulated populations, mortality risk does not vary over time or among individuals, so the observed state (dead or alive) of a collared animal $i$ in month $m$ and year $t$ depends on the true survival rate $\dot{S}_t$ (eq X) as follows: 
$$\hat{I}_{i,t,m} \sim \text{Bernoulli}(\dot{S}^{1/12}_t\hat{I}_{i,t,m-1}).$$

### Observed recruitment from composition surveys

@@

All regression coefficients are assumed to be Gaussian distributed, with priors derived from the expected values and 95% confidence intervals estimated by Johnson et al (2020) (Table \@ref(tab:priors)). We calibrated the prior distributions so that the 95% prior prediction intervals for survival and recruitment from the Bayesian model match the range between the 2.5% and 97.5% quantiles of 1000 simulated survival and recruitment trajectories (Appendix A, Figure \@ref(fig:examplePredictions)).

@@

```{r examplePredictions, fig.cap="Prior (top row) and posterior (bottom row) predictions from the Bayesian model for an example population trajectory. In this example scenario anthropogenic disturbance increases by 2\\% per year from 0 to 100\\%. The population is monitored for 15 years with 30 collars per year, and 6 cows per collared cow in the composition surveys. The prior means and 95\\% predictive intervals (top row, orange lines and shading) are similar to the means and ranges between the 2.5\\% and 97.5\\% quantiles of 3000 simulated survival and recruitment trajectories from the national model (top row, blue lines and shading). Local population data (open triangles) reduces uncertainty (bottom row), but when disturbance is very high local data is not required to predict viability (population growth < 1).", cache=TRUE}

knitr::include_graphics(here::here("analysis/paper/figs/bayesianModelExamples.png"))

@@

TO DO: add example trajectory figure

@@

```{r figStatusErrors, fig.cap="The probability of a correct status assessment varies with the amount of anthropogenic disturbance (columns), the number of years projected (rows), and monitoring effort (see Figure 1 for scenarios). Probabilities are calculated for 200 example population trajectories (see example trajectories in Figure 2). For each example population, the status assessment is correct if the the Bayesian posterior mean and the true population growth rate are both less than or greater than one. When disturbance is very high the probability of correct status assessment is high even without local monitoring. Where disturbance is moderate, increasing the number of years of monitoring and the number of collars reduces uncertainty, but there are diminishing returns on additional monitoring as sample size increases.", cache=TRUE}
knitr::include_graphics(here::here("figs/forPaper/powercmult6ay3aSf1.png"))
```

@@

cr=expand.grid(w = c(1,2,3,4,5,6,7,8,9),rep=seq(1,10000))
nr=nrow(cr)
cr$c = compositionBiasCorrection(q=runif(nr,0,0.6),w=cr$w,u=runif(nr,0,0.2),z=runif(nr,0,0.2))
cs = compositionBiasCorrection(q=runif(nr,0,0.6),w=cr$w,u=runif(nr,0,0.2),z=runif(nr,0,0.2),approx=T)

cr = merge(cr,cs)
cr$ca= exp(rnorm(nrow(cr),mean=cr$mu,sd=cr$sig2^0.5))
